Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences \\ \large Supplementary Information

### 2017-05-16 manuscript/response_to_reviewers.txt

3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

*Response:
The Phylotyper approach is not likely not applicable to core-genome or whole-genome MLST schemes. The value that Phylotyper provides is that it can take as input gene sequences and return predictions for subtypes based on phenotype or subtypes that cannot be directly inferred from the gene sequences alone. For example, PCR or restriction digestion-based subtyping schemes are an abstraction of the underlying gene sequences and in most cases, cannot be directly inferred from an in silico genome sequence. We have demonstrated the ability of Phylotyper to accurately predict a PCR-based subtype (Stx1 and Stx2) as well as a phenotype-based scheme (O-type:H-type serotype).

Subtypes that are directly identified from the genome sequence do not require Phylotyper's additional prediction step. In MLST, the combination each unique allele sequence defines the type.  Potentially, and this is perhaps what the reviewer is refering to, a specific region of the genome that is smaller than the sequence regions used in an MLST, exists that can be used as a proxy to predict MLST subtype. The benefit in this case, is that the entire genome would not need to be sequenced (or multiple alleles interogated) in order to predict MLST type. There is precedent for this type of approach.  The spa and coa genes in Staphylococcus aureus are used as gene markers to infer phylogenetic and population variation.  The spa and coa types have noted concordance with MLST typing. While this is in interesting potential new application for Phylotyper, it is outside the scope of what we originally designed Phylotyper for.

Horizontal gene transfer or recombination would only impact the result of Phylotyper in situations were linkage disequlibrium can occur. If the sequences used to build the subtype scheme produce gene products that determine the subtype, such as the O-antigen processing genes for E. coli being used to predict O-group serotypes, then the prediction would directly be linked to underlying genome. If however a proxy sequence is used, HGT could impact. We have recommended to users building subtype schemes they focus on

### 2017-05-16 manuscript/phylotyper_revised.tex
UPDATE: intro UPDATE: implementation UPDATE: suppl info UPDATE: response to reviewers

Whole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
One of the workhorse methods in public health is molecular subtyping (such as serotyping).
As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.

Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data. 
Phylotyper is unique in that it builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences to help inform subtype prediction. 
Using phylogenetic ancestral state reconstruction to assign the likelihood of each subtype to the tree branch points, Phylotyper assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.

@@

Phylotyper is developed in python and R. 
The steps in the Phylotyper pipeline are: 
(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}. Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded.
(2) Align input genes against a pre-aligned set of reference genes using MAFFT's \texttt{--add} feature \citep{Katoh2013}.
(3) If multiple loci are involved, concatenate individual alignments into superalignment.
(4) Generate maximum likelihood phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
(5) Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
(6) Identify the subtype with maximum marginal likelihood for the unknown genes and report to user in text output file.
Users are also provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).

@@

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequence are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline.
Results are reported to the user.
The first test checks the the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure; F-score, through a leave-one-out cross-validation.
This validation uses each sequence in the training set as a test input (the test sequence is removed from the tree and then used as input).
Results are recorded. 
Metrics like precision, recall and F-score rapidly erode as the phylogenetic grouping is decoupled from the subtypes.
These checks will verify that the sequences phylogenetic grouping is predictive of the subtype subtype.
There is no check that can verify the training set covers all subtypes; this is left up to the user.
However, as we show in the Results section, Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set.

@@

All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test as described earlier.  
The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
The F-score reflects that predictive capability of the subtype scheme. 
If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
The F-score and recall rapidly decrease as the proportion of altered subtypes increased (Supplementary Figure S2).

>>>>

Whole-genome sequencing (WGS) is transforming the public health field by providing an efficient method for surveying bacterial populations.
The speed, discriminatory power and broad utility of WGS can improve surveillance and outbreak analysis.
Adoption of WGS in public health, however, requires transitioning of historical data with the new methods \citep{Jenkins2015}.
One of the workhorse methods in public health is subtyping.
Subtyping methods can broadly categorized as phenotype-based or DNA-based.
Phenotype-based subtypes are, for example, interrogated by biochemical tests (biotyping), detection of surface antigens (serotyping) or susceptibility to bacteriophage (phagetyping) \citep{Wiedmann2002}.
Alternatively DNA-based subtyping examines and classifies bacteria based on genetic content.
DNA-based subtypes use a variety of methods from PCR to Pulse Field Gel Electrophoresis to DNA sequencing to assign groups to bacteria \citep{Wiedmann2002}.
As a surveillance tool, subtypes provide a clearcut designation that is typically used to distinguish taxonomic groups and infer phenotypes, for example, pathogens from non-pathogens.
A WGS-based approach to subtyping would have several benefits over current subtype systems; it would be faster, have improved discrimination and would be cheaper and easier to maintain\citep{Jenkins2015}.
Accordingly, new \textit{in silico} tools have been developed to predict subtypes from WGS data \citep{Joensen2015,Ingle2016,CARRILLO2016}.
To predict from WGS data, these tools target subtypes (Phenotype-based or DNA-based) that can be linked to a specific region or gene in the genome.
These tools have not been applied to subtyping methods that already directly use DNA sequence data to assign subtype such as Multi-Locus Sequnce Typing. 
No inference or extrapolation is needed to integrate a direct DNA sequence-based method, like MLST, into a WGS system.
An example of a subtype system that has been adpated for WGS is serotyping in \textit{Escherichia coli}.
The sequence of O-antigen processing genes in \textit{E. coli} is used to predict O group serotype in \citep{Joensen2015,Ingle2016}.
Another example, is the Shiga toxin subtype (Stx); a DNA-based subtying scheme generated using PCR.
The tool in \citep{CARRILLO2016} predicts Stx subtype by simulating PCR in silico.

Phylotyper is a novel \textit{in silico} predictor of subtypes from sequence data.
Similar to \citep{Joensen2015,Ingle2016,CARRILLO2016}, it also works on subtypes that can be predicted from specific, pre-selected gene or genomic region sequences.
Phylotyper is unique in that it builds a phylogenetic tree consisting of reference sequences with known subtype and the unknown query sequences to help inform subtype prediction.
Using phylogenetic ancestral state reconstruction to assign the likelihood of each subtype to the tree branch points, Phylotyper then assigns an unknown query sequence a subtype based on the extrapolated value from its ancestors in the tree.

@@

Phylotyper is developed in python and R. 
The steps in the Phylotyper pipeline are: 
(1) Identify subtype gene loci in input genomes using BLAST \citep{Camacho2009}. Hits that do not align over 95\% or have under 90\% sequence identity with reference sequence are discarded. Users are notified if no loci are found in genome.
(2) Align input genes against a pre-aligned set of reference genes using MAFFT's \texttt{--add} feature \citep{Katoh2013}.
(3) If multiple loci are involved, concatenate individual alignments into superalignment.
(4) Generate maximum likelihood phylogenetic tree of aligned genes with FastTree \citep{Price2010}.
(5) Run phytools \texttt{rerootingMethod} using the phylogenetic tree and assigned subtypes \citep{Revell2011}.
(6) Identify the subtype with maximum marginal likelihood for the unknown genes and report to user in text output file.
Users are also provided with an image of the phylogenetic tree overlaid with the likelihood values (e.g. Figure~1\vphantom{\ref{fig:01}}).

Detailed descriptions on how to run Phylotyper and the underlying methods are provided here: \url{https://github.com/superphy/insilico-subtyping}.

@@

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
To add a new subtype scheme for use in Phylotyper, users require a training set of homologous sequences with assigned subtype whose phylogenetic grouping is predictive of the subtype.
Phylotyper assumes that the provided training sequences are 1) homologous, specifically, they are suitable for alignment and phylogenetic reconstruction 2) the sequence phylogeny is correlated with the subtype distribution and 3) the set is representative of the range of sequences that make up a subtype.
Where possible, checks are built-in to the \texttt{new} pipeline to ensure the assumptions of the Phylotyper approach are not violated.
Specifically, each new subtype is subject to two tests in the \texttt{new} pipeline and results are reported to the user.
The first test checks that the distribution of inter-patristic phylogenetic distances between the instances of the same subtype is both smaller and distinct from non-identical subtypes.
This test can identify isolated cases of potentially mislabelled subtype genes that are tightly clustered with other subtypes.
A second check computes the accuracy measure, F-score, through a leave-one-out cross-validation; a procedure which uses each sequence in the training set as a test input to estimate positive and negative prediction rates (Supplementary Methods).
The performance metrics: recall and F-score rapidly decrease as the correspondence of training set phylogeny and subtype distribution decreases (Supplementary Figure)
These checks will verify that the training set sequences' phylogenetic grouping is predictive of the subtype.
There is no check that can confirm the training set covers all subtypes in a particular scheme; it is important that the user monitor and update schemes as gaps are identified.
Phylotyper is designed to return a \texttt{non-significant/undetermined} result when encountering a unknown subtype that has no representative in the training set.
Phylotyper is comparatively more robust than other methods as it generates fewer Type-I errors when encountering novel alleles or subtypes not present in the training set (see Results section for details).

@@

All new or updated subtype schemes added in Phylotyper are subject to a leave-one-out cross-validation test.
The test is part of the \texttt{add} pipeline and is used to estimate the F-score of the subtype scheme. 
The F-score reflects that predictive capability of the subtype scheme. 
If the associated phylogeny for training set gene sequences are not correlated with the subtype distribution, this will be reflected in the F-score. 
To demonstrate this property, we randomly assigned subtypes for increasing proportions of the genes in the training set and computed the F-score with the leave-one-out validation for each proportion level.
The F-score and recall rapidly decrease as the proportion of altered subtypes increased (Supplementary Figure S2).

### 2017-05-16 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: intro UPDATE: implementation UPDATE: suppl info UPDATE: response to reviewers

>>>>

\subsection{Phylotyper Verifications Performed on New Subtype Schemes}

Phylotyper conducts two tests for submitted new subtype schemes. The purpose of the first check is to identify isolated cases of distinct subtypes that are clustered together in the phylogenetic tree. In large-scale cases, descendent clades in the phylogenetic tree with multiple distinct subtypes may indicate a subtype that is not correlated with the phylogeny. In limited cases, it might indicate annotation errors. Phylotyper computes an inter-patritristic distances for all leaves in the subtype phylogenetic tree for the given subtype scheme. It then computes a threshold is equivalent to 0.4 probability that inter-patristic distances equal to or greater then threshold belong to the same-subtype distribution. To estimate the distributions, all inter-patristic distances for nodes with the same subtype and all distances for nodes with different subtypes are fit with R's Mclust package. Subtrees with a max inter-patristic distance less than this threshold that have distinct subtypes are flagged. Users will be notified and an updated subtype scheme will be generated with proposed corrections when this situation is detected.

All new subtype schemes are subject to a leave-one-out cross-validation analysis. You will be notified if the F1-score (a equally weighted average of precision and recall) is below 0.9.
